\ZpL: a p-adic precision package

### 2018-02-02 ZpL/ZpL.tex
Adding introduction, fixing a few typos

\usepackage{amsmath,amssymb}
%\usepackage{amsthmnoproof}
\usepackage{amsthm}
\usepackage{mathrsfs}
%\let\bibsection\relax
%\usepackage{amsrefs}
%\usepackage[usenames,dvipsnames]{color}
\usepackage{stmaryrd}
\usepackage{enumerate}
%\usepackage[algoruled,vlined,english,linesnumbered]{algorithm2e}
%\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
\usepackage{comment}
\usepackage{multirow}
\usepackage{xspace}
\usepackage{tabularx,multicol}
\usepackage{tikz}
\usetikzlibrary{patterns}

\definecolor{input}{HTML}{303060}
\definecolor{output}{HTML}{A00000}
\definecolor{string}{HTML}{A02020}
\definecolor{ring}{HTML}{A020A0}
\definecolor{function}{HTML}{A06020}
\definecolor{constructor}{HTML}{205080}
\definecolor{method}{HTML}{205080}
\definecolor{keyword}{HTML}{008000}

@@

\newcommand{\ph}{\vphantom{$A^A_A$}}

@@

\newcommand{\calV}{\mathcal{V}}
\newcommand{\ttv}{\texttt{v}\xspace}
\newcommand{\ttw}{\texttt{w}\xspace}
\newcommand{\tttmp}{\texttt{tmp}\xspace}
\newcommand{\calU}{\mathcal{U}}
\newcommand{\calK}{\mathcal{K}}

\newcommand{\rem}{\,\%\,}

\newcommand{\sage}{\textsc{SageMath}\xspace}
\newcommand{\ZpCR}{\text{\color{output} \rm \tt ZpCR}\xspace}
\newcommand{\ZpFP}{\text{\color{output} \rm \tt ZpFP}\xspace}
\newcommand{\ZpL}{\text{\color{output} \rm \tt ZpL}\xspace}
\newcommand{\ZpLC}{\text{\color{output} \rm \tt ZpLC}\xspace}
\newcommand{\ZpLF}{\text{\color{output} \rm \tt ZpLF}\xspace}

\newcommand{\card}{\text{Card}\:}
\newcommand{\ind}{\text{ind}}
\newcommand{\coind}{\text{coind}}
\newcommand{\hgt}{\text{hgt}}
\newcommand{\cohgt}{\text{cohgt}}
\newcommand{\inp}{\text{in}}
\newcommand{\out}{\text{out}}

@@

\newcommand{\cIn}{{\color{input} \tt \phantom{Zp}In}:}
\newcommand{\cOut}{{\color{red} \tt \phantom{Z}Out}:}
\newcommand{\cZpCR}{{\color{output} \tt ZpCR}:}
\newcommand{\cZpFP}{{\color{output} \tt ZpFP}:}
\newcommand{\cZpLC}{{\color{output} \tt ZpLC}:}
\newcommand{\cZpLF}{{\color{output} \tt ZpLF}:}

@@

\newtheorem{theo}{Theorem}[section]
\newtheorem{lem}[theo]{Lemma}
\newtheorem{prop}[theo]{Proposition}
\newtheorem{cor}[theo]{Corollary}
\newtheorem{quest}[theo]{Question}
\newtheorem{conj}[theo]{Conjecture}
\theoremstyle{definition}
\newtheorem{rmk}[theo]{Remark}
\newtheorem{ex}[theo]{Example}
\newtheorem{deftn}[theo]{Definition}

@@

\author{Xavier Caruso}
  \affiliation{Universit\'e Rennes 1; \\
  \institution{IRMAR}
  \city{Rennes, France}
  \postcode{35042}
}
\email{xavier.caruso@normalesup.org}
\author{David Roe}
  \affiliation{University of Pittsburg; \\
  \institution{Department of Mathematics}
  \city{Pittsburgh, PA, USA}
  \postcode{15260}
}
\email{roed@pitt.edu}
\author{Tristan Vaccon}
  \affiliation{Universit\'e de Limoges; \\
  \institution{CNRS, XLIM UMR 7252}
  \city{Limoges, France}  
  \postcode{87060}  
  }
\email{tristan.vaccon@unilim.fr}

@@

\keywords{Algorithms, $p$-adic precision, Automatic Differentiation}

\begin{abstract}
We present a new package \ZpL for the Computer Algebra system \sage. It 
implements a sharp tracking of precision on $p$-adic numbers, following 
the theory of ultrametric precision introduced in \cite{padicprec}. The 
underlying algorithms are mostly based on Automatic Differentiation 
techniques. We introduce them, study their complexity and discuss our 
design choices.
We illustrate the benefits of our package (in comparison with previous 
implementations) with a large sample of examples coming from Linear 
Algebra, Commutative Algebra and Differential Equations.
\end{abstract}

@@

\begin{itemize}
\item Interval arithmetic, floating point arithmetic
\item short introduction of the precision lemma and lattices
\item goal: tracking precision using this theory
\item short presentation of the package
\item plan of the article
\end{itemize}

\todo{Convention: in matrices, vectors are represented by row-vector.}

@@

We begin our tour of the features of the \ZpL package
with some basic arithmetic computations.
The first step is the definition of the $p$-adic ring or field.

\smallskip

{\noindent \small
\begin{tabular}{rl}
\cIn
 & {\color{ring}\verb?Z2?}\verb? = Zp(2, print_mode=?{\color{string}\verb?'digits'?}\verb?)? \\
 & {\color{ring}\verb?Q2?}\verb? = Qp(2, print_mode=?{\color{string}\verb?'digits'?}\verb?)? \\
\end{tabular}}

\smallskip

\noindent
It means that we will work in various implementation of the 
ring of $2$-adic integers with $20$ digits of precision (which
is the default).
The rings {\color{ring}\verb?Z3?} and {\color{ring}\verb?Q3?} are defined similarly.

We show the results for \ZpCR%, \ZpFP 
and \ZpLC.
Results for \ZpLF are only shown when they differ from \ZpLC.

@@

We first take some random element $x$.

\smallskip

{\noindent \small
\begin{tabular}{rl}
\cIn
 & \verb?x = ?{\color{function}\verb?random_element?}\verb?(?{\color{ring}\verb?Z3?}\verb?,5); x? \\
\cZpCR
 & \verb?...11111? \\
%\cZpFP
% & \verb?...00000000000000011111? \\
\cZpLC
 & \verb?...11111? \\
\end{tabular}}

\smallskip

Multiplication by $3$ is a shift on the digits, whether it 
is done by addition or multiplication.

\smallskip

{\noindent \small
\begin{tabular}{rl@{\hspace{1.5cm}}rl}
\cIn
 & \verb?3*x? &
\cIn
 & \verb?x + x + x? \\
\cZpCR
 & \verb?...111110? &
\cZpCR
 & \verb? ...11110? \\
%\cZpFP
% & \verb?...000000000000000111110? \\
%\cZpFP
% & \verb?...000000000000000111110? \\
\cZpLC
 & \verb?...111110? &
\cZpLC
 & \verb?...111110? \\
\end{tabular}}

\smallskip

%Something is lost however in \ZpCR when performing $x+2x$.
%
%\smallskip
%
%{\noindent \small
%\begin{tabular}{rl}
%\cIn
% & \verb?y = 2*x; y? \\
%\cZpCR
% & \verb?...22222? \\
%%\cZpFP
%% & \verb?...00000000000000022222? \\
%\cZpLC
% & \verb?...22222? \\
%\cIn
% & \verb?x + y? \\
%\cZpCR
% & \verb?...11110? \\
%%\cZpFP
%% & \verb?...000000000000000111110? \\
%\cZpLC
% & \verb?...111110? \\
%\end{tabular}}
%
%\smallskip

\noindent
\todo{Add a short explanation?}

The same phenomenon occurs for multiplication.
%when comparing the computation of $x^3$ and $x \times x^2$.

\smallskip

{\noindent \small
\begin{tabular}{rl@{\hspace{1.5cm}}rl}
\cIn
 & \verb?x^3? &
\cIn
 & \verb?x * x * x? \\
\cZpCR
 & \verb?...010101? &
\cZpCR
 & \verb? ...10101? \\
%\cZpFP
% & \verb?...00000010100000010101? \\
%\cZpFP
% & \verb?...00000010100000010101? \\
\cZpLC
 & \verb?...010101? &
\cZpLC
 & \verb?...010101? \\
\end{tabular}}

\medskip

\ZpL is also well suited for working with coefficients 
with unbalanced precision.

\smallskip

{\noindent \small
\begin{tabular}{rl}
\cIn
 & \verb?x = ?{\color{function}\verb?random_element?}\verb?(?{\color{ring}\verb?Z2?}\verb?,10)? \\
 & \verb?y = ?{\color{function}\verb?random_element?}\verb?(?{\color{ring}\verb?Z2?}\verb?,5)? \\
\cIn
 & \verb?u, v = x+y, x-y? \\
 & \verb?u, v? \\
\cZpCR
 & \verb?(...10111, ...01111)? \\
\cZpLC
 & \verb?(...10111, ...01111)? \\
\end{tabular}}

\smallskip

\noindent
Now, let us compute $u+v$ and compare it with $2x$ (observe that
they should be equal).

\smallskip

{\noindent \small
\begin{tabular}{rl@{\hphantom{\hspace{0.8cm}}}rl}
\cIn
 & \verb?u + v? &
\cIn
 & \verb?2*x? \\
\cZpCR
 & \verb?...00110? &
\cZpCR
 & \verb?...00110100110? \\
\cZpLC
 & \verb?...00110100110? &
\cZpLC
 & \verb?...00110100110? \\
\end{tabular}}

\smallskip

\noindent
No loss in precision appears in \ZpLC, contrary to \ZpCR.
%We remark that \ZpFP also provides all the correct digits.

@@

In linear algebra, determinant and characteristic polynomial are 
notoriously hard to compute: see \cite{caruso-roe-vaccon:15, 
caruso-roe-vaccon:17}. We illustrate this with the following example 
(using a division-free algorithm for the computation of the 
characteristic polynomial) :

\smallskip

{\noindent \small
\begin{tabular}{rl}
\cIn
 & \verb?M = ?{\color{function}\verb?random_element?}\verb?(?{\color{constructor}\verb?MatrixSpace?}\verb?(?{\color{ring}\verb?Q2?}\verb?,3), prec=10)? \\
 & \verb?M.charpoly()? \\
\cZpCR
 & \verb?...00000000000000000001*x^3 + ? \\
 & \verb?...1001011.011*x^2 + ...0111.01*x + 0? \\
\cZpLC
 & \verb?...00000000000000000001*x^3 + ? \\
 & \verb?...1001011.011*x^2 + ...11100111.01*x +? \\
 & \verb?...010000010? \\
\end{tabular}}

%\begin{tabular}{rl}
%\cIn
% & \verb?M = ?{\color{function}\verb?random_element?}\verb?(?{\color{constructor}\verb?MatrixSpace?}\verb?(?{\color{ring}\verb?Q2?}\verb?,3), prec=10)? \\
% & \verb?M.determinant()? \\
%\cZpCR
% & \verb?...000100000100? \\
%\cZpFP
% & \verb? ...11010101010? \\
%\cZpLC
% & \verb?  ...0110110001? \\
%\cZpLF
% & \verb?...001110111100? \\
%\cIn
% & \verb?M.charpoly()? \\
%\cZpCR
% & \verb?(...0000000001)*x^3 + (...1001111111)*x^2 ?\\
% & \verb?+ (...01110101110)*x + (...111011111100)? \\
%\cZpFP
% & \verb?...0000000001*x^3 + ...1101001001*x^2 + ?\\
% & \verb?...0100110011*x + ...00101010110? \\
%\cZpLC
% & \verb?...0000000001*x^3 + ...1011000111*x^2 + ?\\
% & \verb?...0011101000*x + ...1001001111? \\
%\cZpLF
% & \verb?...0000000001*x^3 + ...1111000101*x^2 + ?\\
% & \verb?...0011111111*x + ...110001000100? \\
%\end{tabular}

@@

Our package can be applied to computation over $p$-adic polynomials. A 
natural example is that of the computation of GCD, whose stability has 
been studied in \cite{caruso-2017}.

A naive implementation of the Euclide algorithm
can produce different behaviour depending 
on the type of implementation of the field
of $p$-adic coefficients.

\smallskip

{\noindent \small
\begin{tabular}{rl}
\cIn
 & \verb?S.<x> = ?{\color{constructor}\verb?PolynomialRing?}\verb?(?{\color{ring}\verb?Z2?}\verb?)? \\
 & \verb?P = ?{\color{function}\verb?random_element?}\verb?(S, degree=10, prec=5)? \\
 & \verb?Q = ?{\color{function}\verb?random_element?}\verb?(S, degree=10, prec=5)? \\
 & \verb?D = x^5 + ?{\color{function}\verb?random_element?}\verb?(S, degree=4, prec=8); D? \\
\cZpCR
 & \verb?...00000000000000000001*x^5 + ...11111010*x^4 +? \\ 
 & \verb?...10000000*x^3 + ...11001111*x^2 +? \\
 & \verb?...10000110*x + ...11100010? \\
\cZpLC
 & \verb?...00000000000000000001*x^5 + ...11111010*x^4 +? \\ 
 & \verb?...10000000*x^3 + ...11001111*x^2 +? \\
 & \verb?...10000110*x + ...11100010? \\
\cIn
 & {\color{keyword}\verb?def?}\verb? ?{\color{function}\verb?euclidean?}\verb?(A,B):? \\
 & \verb?    ?{\color{keyword}\verb?while?}\verb? B != 0:? \\
 & \verb?        A, B = B, A % B? \\
 & \verb?    ?{\color{keyword}\verb?return?}\verb? A.monic()? \\
 & \verb??{\color{function}\verb?euclidean?}\verb?(D*P, D*Q)? \\
\cZpCR
 & \verb?0*x^9 + ...1*x^8 + 0*x^7 + 0*x^6 + 0*x^5 +? \\
 & \verb?0*x^4 + 0*x^3 + ...1*x^2 + ...10*x + ...10? \\
\cZpLC
 & \verb?...00000000000000000001*x^5 + ...11111010*x^4 +? \\ 
 & \verb?...10000000*x^3 + ...11001111*x^2 +? \\
 & \verb?...10000110*x + ...11100010? \\
\end{tabular}}

\smallskip

\noindent
With high probability, $P$ and $Q$ are coprime, implying that the gcd of 
$DP$ is $DQ$ is $D$.  However, we observe that we do not always get this 
expected result. More precisely, \ZpCR often founds a gcd of higher 
degree. Indeed the Euclidean algorithm stops prematurely because the 
software believes that some early remainder vanishes due to the lack of 
precision.
%The situation with ZpFP is the exact opposite:
%since the precision is too high (always the maximal cap), 
%the software sometimes does not notice that the remainder vanishes.
%Hence, it happens that the Euclidean algorithm 
%obtains a last non-zero remainder of degree less than that of
%$D.$

@@

In this Section, we explain how our package \ZpL works and we analyze
its performances.
The main theoretical result on which our package is based in the 
ultrametric precision theory developed in \cite{padicprec}, which 
suggests to track precision \emph{via} lattices and differential 
computations. For this reason, our approach is very inspired by 
Automatic Differentiation techniques \cite{} and our implementation 
follows the usual Operator Overloading strategy \cite{}.
We will introduce two versions of our package, namely \ZpLC and \ZpLF: 
this former is safer while the latter is faster.

This Section is organized as follows.
\S \ref{ssec:preclemma} contains a brief overview on the theory of
$p$-adic precision of \cite{padicprec}. 
In the next two subsections, we explain in more details how \ZpLC and 
\ZpLF work. \S \ref{ssec:trackprec} is devoted to the implementation of 
Automatic Differentiation leading to the actual computation of a
global object modeling the precision (which is the aforementioned 
lattice). 
In \S \ref{ssec:viewprec}, we explain how precision on any individual number
can be recoved from the above computations and discuss the validity
of our results.
The overhead complexity induced by our package is analyzed in
\S \ref{ssec:complexity}.

\smallskip

\noindent \textit{Remark about the naming.}
%
The letter \texttt{L}, which appears in the name of the package, 
comes from ``lattices''. The letters \texttt{C} (in \ZpLC) and 
\texttt{F} (in \ZpLF) stand for ``cap'' and ``float'' respectively.

\subsection{The precision Lemma}
\label{ssec:preclemma}

In \cite{caruso-roe-vaccon:14a}, we have suggested the 
use of lattices to represent the precision of elements in 
$K$-vector spaces.  It contrasts with the
\emph{coordinate-wise method} (of \textit{e.g.}
 \verb?Zp(5)?) that is used traditionally in \sage and
where the precision of an element is specified by giving the precision
of each coordinate separately and is updated after each basic
operation.

Consider a finite 
dimensional normed vector space $E$ 
defined over $K$. 
We use the notation $\Vert \cdot \Vert_E$ for the norm 
on $E$ and $B^-_E(r)$ (resp. $B^{\phantom -}_E(r)$) for the open (resp. 
closed) ball of radius $r$ centered at the origin. A \emph{lattice} $L \subset 
E$ is a sub-$\O_K$-module which generates $E$ over $K$. 
Because of ultrametricity, the balls $B^{\phantom 
-}_E(r)$ and $B^-_E(r)$ are examples of lattices. 
Lattices can be thought of as special neighborhoods of $0,$
 and therefore are good 
candidates to model precision data. Moreover, as revealed in 
\cite{caruso-roe-vaccon:14a}, they behave quite well under (strictly) 
differentiable maps:

\begin{prop}
\label{prop:precision}
Let $E$ and $F$ be two finite dimensional normed vector spaces over $K$ 
and $f : U \rightarrow F$ be a function defined on an open subset $U$ of 
$E$. We assume that $f$ is differentiable at some point $v_0 \in U$ and 
that the differential $df_{v_0}$ is surjective.
Then, for all $\rho \in (0, 1]$, there exists a positive real
number $\delta$ such that, for all $r \in (0, \delta)$, any lattice
$H$ such that $B^-_E(\rho r) \subset H \subset B^{\phantom -}_E(r)$ 
satisfies:
\begin{equation}
\label{eq:firstorder}
f(v_0 + H) = f(v_0) + df_{v_0} (H).
\end{equation}
\end{prop}

This proposition enables the \emph{lattice method} of tracking precision,
where the precision of the input is specified as a lattice $H$ and precision
is tracked via differentials of the steps within a given algorithm.
The equality sign in Eq.~\eqref{eq:firstorder} shows that this method
yields the optimum possible precision. 
We refer to \cite[\S 4.1]{caruso-roe-vaccon:14a} for a more complete 
exposition.
We emphasize that polynomials, rational fractions (outside of their poles)
and converging power series are strictly differentiable.

%The lemma is more precise in the context of so-called
%integral polynomial functions.
%A function $f : E \to F$ is said to be
%\emph{integral polynomial} if it is given by multivariate polynomial 
%functions with coefficients in $\O_K$ in any (equivalently all) system 
%of coordinates associated to a $\OK$-basis of $B_E(1)$.
%
%\begin{prop}
%\label{prop:precision2}
%We keep the notation of Proposition \ref{prop:precision} and assume 
%in addition that $f$ is integral polynomial. Let $C$ be a positive real
%number such that $B_F(1) \subset df_{v_0}(B_E(C))$. 
%Then Proposition \ref{prop:precision} holds with $\delta = C \cdot
%\rho^{-1}$.
%\end{prop}

\subsection{Tracking precision}
\label{ssec:trackprec}

We now explain in more details the internal mecanisms \ZpLC and \ZpLF 
use for tracking precision.

In what follows, it will be convenient to use a notion of discrete time 
represented by the letter $t$. Rigorously, it is defined as follows: 
$t=0$ when the $p$-adic ring $\ZpLC(\:\cdots)$ is created and increases by 
$1$ each time a variable is created, deleted\footnote{The deletion can
be explicit (through a call to the \texttt{del} operator) or implicit
(handled by the garbage collector).} or updated.

Let $\calV_t$ be the set of alive variables at time $t$. Set $E_t = 
\Qp^{\calV_t}$; it is a finite dimensional vector space over $\Qp$ which 
should be thought as the set of all possible values that can be taken by 
the variables in $\calV_t$. For $\ttv \in \calV_t$, let $e_\ttv \in
\Qp^{\calV_t}$ be the vector whose all coordinates vanished except 
that at position $\ttv$ which takes the value $1$. The family 
$(e_\ttv)_{\ttv \in \calV_t}$ is obviously a basis of $E_t$; we will 
refer to it as the \emph{canonical basis}.

\subsubsection{The case of \ZpLC.}

Following Proposition~\ref{prop:precision},
the package \ZpLC follows the precision by keeping track of a lattice
$H_t$ in $E_t$, which is a global object whose purpose is to model the 
precision on all the variables in $\calV_t$ all together.
Concretely, this lattice is represented by a matrix $M_t$ in row-echelon 
form whose rows form a set of generators.
Below, we explain how the matrices $M_t$ are updated each time
$t$ increases.

\smallskip

\noindent \textit{Creating a variable.}
%
This happens when we encounter an instruction having one of the
two following forms:

\medskip

\noindent \hspace{5mm} \makebox[2.5cm]{[Computation]\hfill\null}
\verb?w = ?$f$\verb?(v_1,? \ldots\verb?, v_n)?

\smallskip

\noindent \hspace{5mm} \makebox[2.5cm]{[New value]\hfill\null}
\verb?w = ?\verb?R(?\text{value}\verb?, ?\text{prec}\verb?)?

\medskip

\noindent
In both cases, $\ttw$ is the newly created variable. 
The $\ttv_i$'s stand for already defined variables and $f$ is some 
$n$-ary builtin function (in most cases it is just addition, 
subtraction, multiplication or division). On the contrary, the terms 
``value'' and ``prec'' refer to user-specified constants or integral 
values which was computed earlier.

\smallskip

Let us first examine the first construction [Computation].
With our conventions, if $t$ is the time just before the execution of 
the instruction we are interested in, the $\ttv_i$'s lie in $\calV_t$ 
while $\ttw$ does not. Moreover $\calV_{t+1} = \calV_t \sqcup \{\ttw\}$, 
so that $E_{t+1} = E_t \oplus \Qp e_\ttw$.
The mapping taking the values of variables at time $t$ to that at 
time $t{+}1$ is:
$$\begin{array}{rcl}
F: \quad E_t & \longrightarrow & E_{t+1} \smallskip \\
\underline x & \mapsto & \underline x \:\oplus f(x_1, \ldots, x_n)
\end{array}$$
where $x_i$ is the $\ttv_i$-th coordinate of the vector $\underline x$.
The Jacobian matrix of $F$ at $\underline x$ is easily computed; it 
is the block matrix $J_{\underline x}(F) = \big( \begin{matrix} I & L
\end{matrix} \big)$ where $I$ is the identity matrix of size $\text{Card}
\:\calV_t$ and $L$ is the column vector whose $\ttv$-th entry is
$\frac{\partial f}{\partial \ttv} (\underline x)$ if $\ttv$ is one of 
the $\ttv_i$'s and $0$ otherwise.
Therefore, the image of $H_t$ under $dF_{\underline x}$ is represented 
by the matrix $J_{\underline x}(F) \cdot M_t = \big( \begin{matrix} M_t & C
\end{matrix} \big)$ where $C$ is the column vector:
\begin{equation}
\label{eq:defC}
C = \sum_{i=1}^n \frac{\partial f}{\partial{\ttv_i}} (\underline x)
\cdot C_i
\end{equation}
where $C_i$ is the column vector of $M_t$ corresponding to the
variable $\ttv_i$.
Observe that the matrix $J_{\underline x}(F) \cdot M_t$ is no longer a 
square matrix; it has one extra column. This reflects the fact that 
$\dim E_{t+1} = \dim E_t + 1$. Rephasing this in a different language, 
the image of $H_t$ under $dF_{\underline x}$ is no longer a lattice in 
$E_{t+1}$ but is included in an hyperplane.

>>>>

\usepackage{amsmath,amssymb}
%\usepackage{amsthmnoproof}
\usepackage{amsthm}
\usepackage{mathrsfs}
%\let\bibsection\relax
%\usepackage{amsrefs}
%\usepackage[usenames,dvipsnames]{color}
\usepackage{stmaryrd}
\usepackage{enumerate}
%\usepackage[algoruled,vlined,english,linesnumbered]{algorithm2e}
%\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
\usepackage{comment}
\usepackage{multirow}
\usepackage{xspace}
\usepackage{tabularx,multicol}
%\usepackage{tikz}

@@

@@

\newcommand{\ZpL}{\texttt{ZpL}\xspace}
\newcommand{\ZpLCA}{\texttt{ZpLCA}\xspace}
\newcommand{\ZpLCR}{\texttt{ZpLCR}\xspace}
\newcommand{\ZpLF}{\texttt{ZpLF}\xspace}

@@

@@

\newtheorem{theo}{Theorem}[section]
\newtheorem{lem}[theo]{Lemma}
\newtheorem{prop}[theo]{Proposition}
\newtheorem{cor}[theo]{Corollary}
\newtheorem{quest}[theo]{Question}
\newtheorem{conj}[theo]{Conjecture}
\theoremstyle{definition}
\newtheorem{rem}[theo]{Remark}
\newtheorem{ex}[theo]{Example}
\newtheorem{deftn}[theo]{Definition}

@@

\author{Xavier Caruso}
  \affiliation{Universit\'e Rennes 1; \\
  %\institution{IRMAR}
  %\city{Rennes, France}
  %\postcode{35042}
}
\email{xavier.caruso@normalesup.org}
\author{David Roe}
  \affiliation{MIT; \\
  %\institution{Department of Mathematics}
  %\city{Pittsburgh, PA, USA}
  %\postcode{15260}
}
\email{roed@mit.edu}
\author{Tristan Vaccon}
  \affiliation{Universit\'e de Limoges; \\
  %\institution{CNRS, XLIM UMR 7252}
  %\city{Limoges, France}  
  %\postcode{87060}  
  }
\email{tristan.vaccon@unilim.fr}

@@

\keywords{Algorithms, $p$-adic precision, characteristic polynomial,
eigenvalue}

\begin{abstract}
\end{abstract}

@@

When computing with real and $p$-adic fields, exact results are usually impossible, since most elements have infinite decimal or $p$-adic expansions.  Working with these fields thus requires an analysis of how precision evolves through the sequence of steps involved in carrying out a computation.  In this paper, we describe a package for computing with $p$-adic rings and fields, based on a series of papers by the same authors \todo{include citations}.  The core of the package is a method for tracking precision using $p$-adic lattices which can yield dramatically more precise results, at the cost of increased runtime and memory usage.

The standard method for handling precision when computing with real numbers is floating point arithmetic, which may also be used in $p$-adic computation.  At a given precision level, a finite set of representable numbers are chosen, and arithmetic operations are defined to give a representable number that is close to the true result \todo{add reference to floating point arithmetic}.  For example, the representable numbers in $53$-bit floating point arithmetic are those of the form $m \cdot 2^e$ \todo{describe valid $m$ and $e$}.  Floating point arithmetic has the benefit of efficient arithmetic operations, but users are responsible for tracking the precision of the results.  Numerically unstable algorithms can lead to very inaccurate answers \todo{add reference for numerical stability}.

If provably correct results are desired, interval arithmetic provides an alternative to floating point.  Instead of just tracking an approximation to the answer, the package also tracks a radius within which the true result lies.  This method is commonly used for $p$-adic computations since the ultrametric property of $p$-adic fields frequently keeps the radius small.  Computations remain fairly efficient with this approach, but numerical instability can still lead to dramatic losses in precision \todo{include reference to SOMOS section}.

For better control of precision, we may use ellipsoids rather than balls.  This change would have little utility for real numbers, since ellipsoids are not preserved by most functions.  For $p$-adic fields, in contrast, sufficiently small ellipsoids are preserved by differentiable maps.  From an algebraic perspective, these ellipsoids are just cosets of a lattice $H$ inside a $p$-adic vector space, and Lemma \todo{add number} describes how the image of such a coset under a map $f$ is given exactly by applying the differential of $f$ to $H$.

In this paper, we describe an implementation of this idea in Sage \todo{cite Sage}.  \todo{Add short description of package features.}

\todo{Plan of article}

@@

@@

\begin{tabular}{|l|l|}
\verb?R = Zp(3,5)? & \verb?R = ZpL(3,5)? \\
\verb?x = R(143,5); x? & \verb?x = R(143,5); x? \\
\hfill\verb?...12022? & \hfill\verb?...12022? \\
\verb?3*x? & \verb?3*x? \\
\hfill\verb?...120220? & \hfill\verb?...120220? \\
\verb?x + x + x? & \verb?x + x + x? \\
\hfill\verb?...20220? & \hfill\verb?...120220? \\
\verb?x^3? & \verb?x^3? \\
\hfill\verb?...020222? & \hfill\verb?...020222? \\
\verb?x * x * x? & \verb?x * x * x? \\
\hfill\verb?...20222? & \hfill\verb?...020222? \\
\end{tabular}

\begin{tabular}{|l|l|}
\verb?R = Zp(2,10)? & \verb?R = ZpL(2,10)? \\
\verb?x = R(987,10); x? & \verb?x = R(987,10); x? \\
\hfill\verb?...1111011011? & \hfill\verb?...1111011011? \\
\verb?y = R(21,5); y? & \verb?y = R(21,5); y? \\
\hfill\verb?...10101? & \hfill\verb?...10101? \\
\verb?u = x + y; u? & \verb?u = x + y; u? \\
\hfill\verb?...10000? & \hfill\verb?...10000? \\
\verb?v = x - y; v? & \verb?v = x - y; v? \\
\hfill\verb?...00110? & \hfill\verb?...00110? \\
\verb?u + v? & \verb?u + v? \\
\hfill\verb?...10110? & \hfill\verb?...11110110110? \\
\end{tabular}

@@

\begin{itemize}
\item determinant
\item characteristic polynomial
\end{itemize}

@@

\begin{itemize}
\item gcd of polynomials
\end{itemize}

\section{Theory}

\subsection{The precision Lemma}

\subsection{The way the precision is tracked with \ZpL}

We follow the precision lattice.
Closely related to automatic differentiation.

\begin{itemize}
\item \ZpLCR, \ZpLCA: we use a cap to ensure that we have a lattice at
each step.
\item \ZpLF: we work with $\Zp$-submodules of possible positive codimension
and represent them using matrices of floating point $p$-adics.
\end{itemize}

\subsection{The notion of working precision}

\begin{itemize}
\item what is it?
\item how can we compute it? (Answer: Hermite reduction of the transpose)
\item how can we estimate it?
\end{itemize}

\subsection{Correctness}

\begin{itemize}
\item prove that \ZpLCA and \ZpLCR are correct under some assumptions
(to be determined)
\item explain that \ZpLF is not meant to provide proved results
\end{itemize}

\subsection{Complexity}

\begin{tabular}{l|c|c|}
& \ZpLCA/\ZpLCR & \ZpLF \\
\hline
Creation of a variable & $O(n)$ & $O(n_0)$ \\
\hline
Deletion of a variable & $O(m^2)$ & $-$ \\
\hline
\end{tabular}

discussion on temporal locality

@@

\begin{itemize}
\item Discuss coercion/conversion: \texttt{Zp} coerces to \ZpL but no coersion
in the other direction
\item How to access to the precision
\item Diffused digits
\item Historic of variables creation/deletion, etc...
\item Other features?
\end{itemize}

\subsection{More examples}

\begin{tabular}{|l|l|}
\verb?R = Zp(2)? & \verb?R = ZpL(2)? \\
\multicolumn{2}{|l|}{\tt x, y, z, t = R(1,15), R(1,15), R(1,15), R(3,15)} \\
\multicolumn{2}{|l|}{\tt for \_ in range(20):} \\
\multicolumn{2}{|l|}{\tt \ \ \ \ x, y, z, t = y, z, t, (y*t + z*z)/x} \\
\multicolumn{2}{|l|}{\tt \ \ \ \ print t} \\
\hfill\verb?...000000000000100? & \hfill\verb?...000000000000100? \\
\hfill\verb?...000000000001101? & \hfill\verb?...000000000001101? \\
\hfill\verb?...000000000110111? & \hfill\verb?...000000000110111? \\
\hfill\verb?...101010111010111? & \hfill\verb?...101010111010111? \\
\hfill\verb?...1100111101111? & \hfill\verb?...101100111101111? \\
\hfill\verb?...1110000010010? & \hfill\verb?...111110000010010? \\
\hfill\verb?...0001000111001? & \hfill\verb?...100001000111001? \\
\hfill\verb?...0000011111101? & \hfill\verb?...100000011111101? \\
\hfill\verb?...1000000110101? & \hfill\verb?...001000000110101? \\
\hfill\verb?...101101010011? & \hfill\verb?...010101101010011? \\
\hfill\verb?...110000000000? & \hfill\verb?...001110000000000? \\
\hfill\verb?...000101011101? & \hfill\verb?...111000101011101? \\
\hfill\verb?...001001101011? & \hfill\verb?...111001001101011? \\
\hfill\verb?...000011110011? & \hfill\verb?...111000011110011? \\
\hfill\verb?...11? & \hfill\verb?...0000000111? \\
\hfill\verb?...10? & \hfill\verb?...1010001110? \\
\hfill\verb?...01? & \hfill\verb?...1110110001? \\
\hfill\verb?...01? & \hfill\verb?...011110111001? \\
\hfill\verb?...01? & \hfill\verb?...100101101001001? \\
\hfill\verb?...1? & \hfill\verb?...11111000011? \\
\end{tabular}

\begin{itemize}
\item SOMOS sequence
\item many multiplications of matrices
\item Dodgson algorithm
\item subresultants
\item Grobner basis
\item $p$-adic differential equations
\end{itemize}

\subsubsection{$p$-adic differential equations}

In \cite{LV16}, the behaviour of the precision when solving
$p$-adic differential equations with separation of 
variables has been studied.
The authors have investigated the gap when that appears
when applying a Newton-method solver between
the theoretic loss in precision and the 
actual loss in precision for a naive implemenationin \verb?Zp(p)?.
We can reach this theoretical loss in precision using \verb?ZpL?.
We use a generic \verb?Newton_Iteration_Solver(g,h,N)?
that applies \verb?N? steps of the Newton method for 
$y'=g \times h(y)$ as described in \cite{LV16}.

\begin{tabular}{|l|l|}
\verb?R = Qp(2,250)? & \verb?R = QpL(2,250)? \\
\verb?S = Zp(2,250)? & \verb?S = ZpL(2,250)? \\
\multicolumn{2}{|l|}{\tt A.<x> = PowerSeriesRing(R,32) } \\
\multicolumn{2}{|l|}{\tt B.<x> = PowerSeriesRing(S,32)) } \\
\multicolumn{2}{|l|}{\tt h,y =1+x+x\^{}3 , x+x\^{}2*A(B.random\_element()) } \\
\multicolumn{2}{|l|}{\tt g=y.derivative()/h(y) } \\
\multicolumn{2}{|l|}{\tt s=Newton\_Iteration\_Solver(g,h,5)} \\
\multicolumn{2}{|l|}{\tt print 250-(s[31].lift()-y[31].lift()).valuation(2)} \\
\hfill\verb?9? & \hfill\verb?4? \\
\end{tabular}
The theorical loss in precision for the coefficient of $t^{31}$
is $\lfloor log_2 (31)\rfloor=4$ and is then reached, whereas the
loss in precision for a naive implementation is clearly more severe.
